A comparison of language control while switching within versus between languages in younger and older adults.

### 2023-08-22 manuscriptINSPIRE.Rmd
Load in kableExtra
Update R1 manuscriptINSPIRE.Rmd

devtools::install_github("kupietz/kableExtra")

@@

All analyses included Rule (noun coded as -0.5; verb as 0.5; or size coded as -0.5 and parity as 0.5); Age group (younger adults coded as -0.5; older adults as 0.5); and trial type (switching: non-switch coded as -0.5 and switch trials as 0.5; mixing: single-rule coded as -0.5 and non-switch trials as 0.5). For the accuracy analysis, we excluded trials in the dual-rule condition preceded by a break as they did not have a trial type (switch or non-switch). For RT analyses, we excluded incorrect responses and trials preceded by a break, no response, or wrong-language response. Prior to the RT analyses, we also removed RT outliers that fell 2.5SD above or below the mean by participant and condition (1.6% of correct noun-verb task trials and 1.9% of non-linguistic task trials, [45]). Visual inspection of the RT data showed that they were not normally distributed and we therefore conducted the analysis with log-transformed RTs. Descriptive statistics provided in the text and tables or figures are based on untransformed data.

@@

Finally, in addition to these analyses, we also examined whether the mixing and switching costs were similar across the linguistic and non-linguistic switching task. For each participant, we therefore computed their switching cost (RT difference between switch and non-switch trials) and mixing cost (RT difference between non-switch and single-rule trials) per task and conducted correlational analyses comparing the costs across tasks.

@@

The full results of the mixing cost error analysis are provided in Supplementary Table \@ref(tab:TableS2). The error analysis firstly revealed a mixing cost, with participants making fewer errors in the single-rule condition than in the dual-rule condition (see Supplementary Table \@ref(tab:TableS1)). No further effects of, or interactions with, age group or rule were found.

@@

Supplementary Table \@ref(tab:TableS5) provides the full results from the switching- and mixing-effect analyses. Both analyses (see Table \@ref(tab:Table3)) showed that older adults responded more slowly (*M* across all trial types = `r round(NLSswitchM2[1,2])`, *SD* = `r round(NLSswitchSD2[1,2])`) than younger adults (*M* across all trial types = `r round(NLSswitchM2[2,2])`, *SD* = `r round(NLSswitchSD2[2,2])`). Surprisingly, no switching effect was found (see Table 3), reflecting that switch (*M* = `r round(NLSswitchM[3,2])`, *SD* = `r round(NLSswitchSD[3,2])`) and non-switch RTs (*M* = `r round(NLSswitchM[1,2])`, *SD* = `r round(NLSswitchSD[1,2])`) were not significantly different. If anything, people were responding more slowly to switch trials, contrary to the hypothesis. There was a mixing effect, reflecting that responses were fastest on single-rule trials (*M* = `r round(NLSswitchM[2,2])`, *SD* = `r round(NLSswitchSD[2,2])`).

The switching cost was not significantly different between age groups. The mixing cost did differ between younger and older adults. In terms of untransformed RTs, older adults showed slightly higher mixing costs (see Table 3). However, in terms of log RTs (as used in the analysis), older adults’ mixing cost was smaller (*M* cost = `r round(logRT[1],digits=2)`, *SD* = `r round(logSD[1],digits=2)`) than that of younger adults (*M* cost = `r round(logRT[2],digits=2)`, *SD* = `r round(logSD[2],digits=2)`). Given that the direction differed between log RTs and untransformed RTs, and considering the large difference in overall RTs between age groups, we conducted an additional analysis using z-scored RTs. The interaction between age and mixing costs was not significant in this analysis ($\beta$ = `r round (Zscore_NLS$coefficients[c(5),c(1)],digits=3)`, SE = `r round (Zscore_NLS$coefficients[c(5),c(2)],digits=3)`, t = `r round (Zscore_NLS$coefficients[c(5),c(4)],digits=3)`, p = `r p_format (Zscore_NLS$coefficients[c(5),c(5)],digits=3,leading.zero=TRUE,accuracy=0.001)`).

@@

## Discussion
Experiment 1 tested a noun-verb switching task to examine whether markers of language control (switching and mixing costs) frequently observed in bilinguals are observed when switching within a language too. It furthermore compared those effects between younger and older adults, and compared the costs to a non-linguistic switching task. The noun-verb switching task showed both switching and mixing costs, but no effects of age. The non-linguistic switching task showed, surprisingly, only a mixing and no switching cost.

### Within-language noun-verb switching
The noun-verb switching task showed that participants needed more time when switching between rules (switching cost) as well as when using two rules in a dual-rule condition compared to a single-rule condition (mixing cost). The presence of mixing costs suggests that participants recruited additional resources to monitor cues and to select responses according to the rules indicated by the cues. The presence of switching costs furthermore shows that participants experienced competition between the different responses. Interestingly, the noun-verb mixing cost also showed an asymmetry, with participants showing larger mixing costs for the easier rule (object/noun naming). A similar pattern was observed in terms of accuracy too. Although that did not reach significance, it shows the asymmetry was not the effect of a speed-accuracy trade-off for, for example, the noun responses.

This pattern is in the same direction as asymmetries often observed in the bilingual production literature, either regarding switching costs (e.g., [10]) or mixing costs (e.g., [15]). These asymmetries are often explained through a form of top-down control, with people potentially inhibiting the easier responses to allow for flexible use of responses that are more difficult to retrieve. The finding that this asymmetry concerned mixing costs suggests that participants might have proactively inhibited noun responses to more easily use verb responses too. Alternatively, or additionally, they might have proactively over-activated the verb responses (e.g., [46]). This suggests some form of control is involved even in noun-verb switching. If the costs observed were purely due to more difficult words requiring more time to reach their activation threshold, the verbs should have shown larger mixing or switching costs. Instead, this language control over the dominant rule (nouns) might be needed in particular when speakers are required to alternate between responses in high-competition situations such as our dual-rule task.

Interestingly, age did not influence the mixing nor switching costs. More generally, within the dual-rule context, no overall age effect on naming was observed either. This is surprising as overall slower naming in older adults would be expected to be present, or even most pronounced, in the most-difficult dual language context. By comparing age effects on this task with the bilingual switching task in Experiment 2, we aim to better understand whether these groups of older adults performed at a similar level as younger adults regardless of the task or whether an absence of age-group differences is specific to within-language switching.

### Task switching
Experiment 1 also included a colour-shape (non-linguistic) task switching paradigm. Surprisingly, this task did not show a switching cost in terms of RTs and showed a switching benefit in terms of accuracy, contrary to a vast literature observing costs when switching between a range of tasks [47]. A large mixing cost was observed, showing that using different rules was more difficult than being in a single-rule condition. It is possible that the dual-rule condition was very difficult overall, even when not having to switch between rules. Even on non-switch trials, participants had to choose between one of two possible responses corresponding to the rule (for example, smaller or larger than 5 within the size-rule task). Responses in a linguistic noun-verb naming task might be more automatic, as language production is natural and people are fast at retrieving a word like “horse” in response to a picture. In contrast, determining whether 7 is smaller or larger than 5 might be difficult even on non-switch trials and is not a task people frequently perform in daily life. Furthermore, this might have been even more difficult in the online set-up, where participants had to click on, and remember, buttons corresponding to these options.

The very large mixing cost further confirms that the dual-rule context was difficult for both age groups and that participants might have put in much additional effort to respond to both non-switch and switch trials. A small difference between age groups was observed in the analysis on log RTs, although the direction varied for untransformed and log-transformed RTs. While the untransformed RTs suggest older adults had a larger mixing costs, the analysis on log RTs showed a smaller mixing cost for older adults. It is likely that the untransformed RTs show a larger raw mixing cost for older adults as their overall RTs were much slower than that of younger adults. This large increase in overall response times in older adults was only observed in the non-linguistic task, probably because the time out was much longer in this task (ten seconds). We therefore also conducted an additional analysis with z-scored RTs to cancel out the overall age effect. This analysis no longer showed a significant age-group difference in terms of mixing costs. In combination with the benefit older adults showed in terms of the accuracy mixing cost too, it appears that older adults’ responses were not affected more negatively by dual-rule compared to single-rule contexts than younger adults’ responses. Thus, overall, neither the linguistic nor the non-linguistic tasks showed disadvantages for older adults.

@@

As a second aim, we examined how age affects within- and between-language control. Experiment 1 showed no clear language control differences between age groups, but previous literature has shown substantial age effects on bilingual control in terms of both reactive and proactive control mechanisms. If older adults show greater difficulty with proactive control, we expected their mixing costs to be larger than for younger adults. If these age-related changes are comparable across within- and between-language control, we expected these effects to be observed in both tasks, and for the age effects to be comparable. Similarly, if older adults show greater difficulty with reactive control, we expected their switching costs to be larger than those of younger adults, again also examining if age-related patterns are comparable for the within- and between-language control tasks.

@@

### Participants
Experiment 2 was completed by fifty older adults (Mage = 67.20, SD = 5.60, range = 60-82) and fifty younger adults (Mage = 23.1, SD = 4.97, range = 18-35). Participants provided informed consent at the start of the study. Ethics approval was given by the ethics committee in the Department of Psychology, University of York. Thirty-four of the older adults identified as female and 16 as male. Within the group of younger adults, 34 identified as female, 13 as male, and three as one of the other options. Thirty-eight of the older adults and twenty-five of the younger adults reported having completed a graduate degree. An additional seven participants were tested but not included because their audio files were not uploaded or audible (N=5) or because they did not follow the task instructions (N=2). The sample size was determined based on power simulations. We first ran a simulation assessing power to detect the switching and mixing costs observed for the noun-verb task in Experiment 1. With 100 participants, power to detect these effects was over 95%. Given that Experiment 1 showed no age-group differences, we based the power in relation to age groups on the bilingual switching study with older and younger adults reported in [21]. Power to detect a mixing cost difference between age groups was over 80%. Power to detect a switching cost difference between age groups, based on effect sizes in [21], was slightly lower (>60%). However, based on practical considerations regarding recruitment feasibility, we decided to accept this power with 100 participants.

All participants reported that they were native speakers of English and met the same inclusion criteria as reported in Experiment 1, with the exception that participants were bilingual in Experiment 2. Participants spoke French (39 older adults; 34 younger adults) or German (11 older adults; 16 younger adults) as a second language. We recruited both French- and German-speakers to facilitate recruitment but did not aim to compare French and German speakers. Participants completed the LexTALE as a short measure of vocabulary in each language ([48],[49]). In this task, they saw letter strings and had to indicate whether they formed an existing word in the target language or not. Participants also completed a language background questionnaire, which included questions about age of acquisition, self-rated proficiency, language use, and language switching. Participants reported being unbalanced bilinguals, with a higher proficiency in and use of English than their L2 (see Supplementary Table 6). Most participants were living in an English-dominant country (typically the UK). Some participants also reported speaking more than two languages, but for the purpose of our study, we always counted French or German as the L2, depending on the language version of the task they completed. We initially pre-registered to exclude participants who scored below 50% correct on the LexTALE. However, given that we aimed for unbalanced bilinguals with a lower L2 proficiency, several participants scored below this threshold and we removed this exclusion criterion.

@@

### Materials
Pictures for the two switching tasks were chosen in a similar way as in Experiment 1, with the list of stimuli available on the OSF page. We used the same pictures in the bilingual and in the noun-verb switching task to make them more comparable. Identical cognates between the languages were avoided, which also resulted in a different set of stimuli for the noun-verb task than in Experiment 1. For some pictures, multiple verbs were possible (e.g., “chew”, “brush”, or “bite”), which we all accepted. Words were matched in terms of number of phonemes between languages or nouns/verbs, but the verb and L2 frequency was significantly higher than the noun and L1 frequencies. Materials for the semantic control task were based on [7]. Again, two cues per switching task were used. The cues for the noun-verb switching task were the same as in Experiment 1. For the bilingual switching task, we could not create two written cues per language and were therefore worked with two visual cues in the form of two versions of the country flag associated with the language.

### Procedure
Participants were recruited through Prolific.co and through our existing database of older adult volunteers. They completed the study on Gorilla.sc. The procedure of the study was similar to Experiment 1, with the order of the noun-verb and bilingual switching tasks counterbalanced across participants. The tasks looked the same as in Experiment 1, with the change that each picture was presented for 3000ms. After the two switching tasks, participants completed a semantic control task based on [7]. In this task, participants saw a target word and two or four answer options. They had to choose the answer option that was similar in either size (e.g., donkey – bicycle) or colour (e.g., cloud-tooth). Other answer options were either neutral (congruent trials) or distractors (incongruent trials) that were related to the target in another way. For example, on the incongruent trial asking what is most similar to “gym” in size, participants had to choose “theatre” while ignoring the related word “basketball”. Half of the trials were incongruent while the other half were congruent. There was no time limit on this task. After completing these tasks, participants completed the LexTALEtasks as a measure of vocabulary and the language background questionnaire, as described in the Participants section.

@@

***Accuracy*** 
Mean accuracy per condition and language group is provided in Supplementary Table \@ref(tab:TableS7). The full results from the analysis are shown in Supplementary Table \@ref(tab:TableS8). The bilingual switching task showed a significant switching cost in terms of accuracy, with participants making more mistakes on switch than on non-switch trials. They were also more accurate in their L1 than in their L2. Accuracy did not show any effects of, or interactions with, age group.

@@

The full results of the switching-effect analysis can be found in Table \@ref(tab:Table4), with Table \@ref(tab:Table5) showing the means per condition. Participants showed a significant switching cost, with slower responses on switch trials (*M* = `r round(Bil_Trialtype_RT_M[3,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[3,2])`) than on non-switch trials (*M* = `r round(Bil_Trialtype_RT_M[1,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[1,2])`). Older adults (*M* = `r round(Bil_Age_RT_M[1,2])`, *SD* = `r round(Bil_Age_RT_SD[1,2])`) responded more slowly than the younger adults (*M* = `r round(Bil_Age_RT_M[2,2])`, *SD* = `r round(Bil_Age_RT_SD[2,2])`). Of main interest for the current study, older adults had a slightly higher switching cost than younger adults, but this was not significant (see Table \@ref(tab:Table4) and Figure 1. None of the other main effects or interactions were significant either.

@@

knitr::kable(table4data, booktabs = T, caption='Outcome of the linear mixed effect models for the bilingual switching task in Experiment 2, The final switching model included all by-subject and by-item random intercepts and slopes, after removal of correlations. The final mixing model included by-subject and by-item random intercepts and slopes, after removal of correlations.', align='lrrrr') %>%
  row_spec(c(0,1,10),bold=T) %>%
  row_spec(c(1,10),italic=T)
```

@@

#plot nounverb mixing cost
NVMix = ggplot(means_NVMix_Exp2_long, aes(x = Agegroup, y = NounVerbMixing)) + geom_boxplot() +
  labs(title="C. Noun-verb mixing cost",x="Age group", y = "mixing cost (ms)")+
  theme(axis.text.x = element_text(color = "black", size = 12, face = "plain"),
        axis.text.y = element_text(color = "black", size = 12, face = "plain"),
        axis.title.x = element_text(color = "black", size = 14, face = "bold"),
        axis.title.y = element_text(color = "black", size = 14, face = "bold"),
        title = element_text(color = "black", size = 14, face = "bold.italic"), 
        strip.text.x = element_text(color = "black", size = 12, face = "plain"))+

@@

***Accuracy*** 
The full details are provided in Supplementary Tables \@ref(tab:TableS7) and \@ref(tab:TableS8). Participants showed a significant mixing cost, with higher accuracy on single-language trials than on dual-language trials. They were also more accurate in their L1 than L2. The mixing cost itself was also larger for the L1. The difference between languages (L1 naming advantage) was more pronounced on single-language than dual-language trials. Finally, this asymmetry in mixing cost interacted with age, as the difference between languages was largest for the younger adults.

***Reaction times***
Participants showed a significant RT mixing cost, with slower responses on non-switch trials (*M* = `r round(Bil_Trialtype_RT_M[1,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[1,2])`) than on single-rule trials (*M* = `r round(Bil_Trialtype_RT_M[2,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[2,2])`). Across the single-rule and non-switch trials, older adults (M = 1229, SD = 167) also responded more slowly than younger adults (M = 1136, SD = 145). Of main interest for the current study, mixing costs were larger in the older adults than in the younger adults (see Table 5 & Figure 1).

@@

Given that older adults also responded significantly slower overall than younger adults, we z-scored the RTs to account for overall slowing. These analyses aimed to make sure that the larger mixing costs (and to some extent larger switching costs) for older adults were not the consequence of their overall slower naming. These analyses still showed a significant interaction between mixing costs and age ($\beta$ = `r round (MixingZ$coefficients[c(5),c(1)],digits=3)`, SE = `r round (MixingZ$coefficients[c(5),c(2)],digits=3)`, t = `r round (MixingZ$coefficients[c(5),c(4)],digits=3)`, p = `r p_format (MixingZ$coefficients[c(5),c(5)],digits=1,leading.zero=TRUE,accuracy=0.001)`), and the interaction between switching cost and age was now also significant ($\beta$ = `r round (SwitchingZ$coefficients[c(5),c(1)],digits=3)`, SE = `r round (SwitchingZ$coefficients[c(5),c(2)],digits=3)`, t = `r round (SwitchingZ$coefficients[c(5),c(4)],digits=3)`, p = `r p_format (SwitchingZ$coefficients[c(5),c(5)],digits=1,leading.zero=TRUE,accuracy=0.001)`).

@@

***Accuracy*** 
Mean accuracy by condition is shown in Supplementary Table \@ref(tab:tableS9), with full results from the analysis shown in Supplementary Table \@ref(tab:TableS10). There was a switch cost in terms of accuracy, with better performance on non-switch than switch trials. This interacted with rule, such that the switching cost was larger for the verbs. There were no effects of, or interactions with, age.

>>>>

devtools::install_github("kupietz/kableExtra")
library(kableExtra)

@@

All analyses included Rule (noun coded as -0.5; verb as 0.5; or size coded as -0.5 and parity as 0.5); Age group (younger adults coded as -0.5; older adults as 0.5); and trial type (switching: non-switch coded as -0.5 and switch trials as 0.5; mixing: single-rule coded as -0.5 and non-switch trials as 0.5). For the accuracy analysis, we excluded trials in the dual-rule condition preceded by a break as they did not have a trial type (switch or non-switch). For RT analyses, we excluded incorrect responses and trials preceded by a break, no response, or wrong-language response. Prior to the RT analyses, we also removed RT outliers that fell 2.5*SD* above or below the mean by participant and condition (1.6% of correct noun-verb task trials and 1.9% of non-linguistic task trials, [50]). Visual inspection of the RT data showed that they were not normally distributed and we therefore conducted the analysis with log-transformed RTs. Descriptive statistics provided in the text and tables or figures are based on untransformed data.

@@

Finally, in addition to these analyses, we also examined whether the mixing and switching costs were related across the linguistic and non-linguistic switching tasks. For each participant, we therefore computed their switching cost (RT difference between switch and non-switch trials) and mixing cost (RT difference between non-switch and single-rule trials) per task and conducted correlational analyses comparing the costs across tasks.

@@

The full results of the mixing cost error analysis are provided in Supplementary Table \@ref(tab:TableS2). The error analysis revealed a mixing cost, with participants making fewer errors in the single-rule condition than in the dual-rule condition (see Supplementary Table \@ref(tab:TableS1)). No further effects of, or interactions with, age group or rule were found.

@@

Supplementary Table \@ref(tab:TableS5) provides the full results from the switching- and mixing-effect analyses. Both analyses (see Table \@ref(tab:Table3)) showed that older adults responded more slowly (*M* across all trial types = `r round(NLSswitchM2[1,2])`, *SD* = `r round(NLSswitchSD2[1,2])`) than younger adults (*M* across all trial types = `r round(NLSswitchM2[2,2])`, *SD* = `r round(NLSswitchSD2[2,2])`). Surprisingly, no switching effect was found (see Table 3), reflecting that switch (*M* = `r round(NLSswitchM[3,2])`, *SD* = `r round(NLSswitchSD[3,2])`) and non-switch RTs (*M* = `r round(NLSswitchM[1,2])`, *SD* = `r round(NLSswitchSD[1,2])`) were not significantly different. If anything, people were responding more slowly to non-switch trials, contrary to the hypothesis. There was a mixing effect, reflecting that responses were fastest on single-rule trials (*M* = `r round(NLSswitchM[2,2])`, *SD* = `r round(NLSswitchSD[2,2])`).

The switching effect was not significantly different between age groups. The mixing cost did differ between younger and older adults. In terms of untransformed RTs, older adults showed slightly higher mixing costs (see Table 3). However, in terms of log RTs (as used in the analysis), older adults’ mixing cost was smaller (*M* cost = `r round(logRT[1],digits=2)`, *SD* = `r round(logSD[1],digits=2)`) than that of younger adults (*M* cost = `r round(logRT[2],digits=2)`, *SD* = `r round(logSD[2],digits=2)`). Given that the direction differed between log RTs and untransformed RTs, and considering the large difference in overall RTs between age groups, we conducted an additional analysis using z-scored RTs. The interaction between age and mixing costs was not significant in this analysis ($\beta$ = `r round (Zscore_NLS$coefficients[c(5),c(1)],digits=3)`, *SE* = `r round (Zscore_NLS$coefficients[c(5),c(2)],digits=3)`, *t* = `r round (Zscore_NLS$coefficients[c(5),c(4)],digits=3)`, *p* = `r p_format (Zscore_NLS$coefficients[c(5),c(5)],digits=3,leading.zero=TRUE,accuracy=0.001)`).

@@

## Discussion
Experiment 1 tested a noun-verb switching task to examine whether markers of language control (switching and mixing costs) frequently observed in bilinguals are observed when switching within a language too. It furthermore compared those effects between younger and older adults, and compared the costs to a non-linguistic switching task. The noun-verb switching task showed both switching and mixing costs, but no age-group differences. The non-linguistic switching task showed, surprisingly, only a mixing and no switching cost.

### Within-language noun-verb switching
The noun-verb switching task showed that participants needed more time when switching between rules (switching cost) as well as when using two rules in a dual-rule condition compared to a single-rule condition (mixing cost), in line with [33]. The presence of mixing costs suggests that participants recruited additional resources to monitor cues and to select responses according to the rules indicated by the cues. The presence of switching costs furthermore shows that participants experienced competition between the different responses. Interestingly, the noun-verb mixing cost also showed an asymmetry, with participants showing larger RT mixing costs for the easier rule (object/noun naming). Although this did not reach significance, a similar numerical pattern was present in the accuracy analysis too, showing that the RT asymmetry was not the consequence of a speed-accuracy trade-off for, for example, the noun responses.

This pattern is in the same direction as asymmetries often observed in the bilingual production literature, either regarding switching costs (e.g., [10]) or mixing costs (e.g., [15]). These asymmetries are often explained through a form of language control, with people potentially inhibiting the easier responses to allow for flexible use of responses that are more difficult to retrieve. The finding that this asymmetry concerned mixing costs suggests that participants might have proactively inhibited noun responses to more easily use verb responses too. Alternatively, or additionally, they might have proactively over-activated the verb responses (e.g., [17]). This suggests some form of control is involved even in noun-verb switching. If the costs observed were purely due to more difficult words requiring more time to reach their activation threshold, the verbs should have shown larger mixing or switching costs. Instead, monolingual production might use language control too, in particular when speakers are required to alternate between responses in high-competition situations such as our dual-rule task.

Interestingly, age did not influence the mixing nor switching costs. More generally, within the dual-rule context, no overall age-group difference in terms of overall naming RTs was observed either. This is surprising as overall slower naming in older adults would be expected to be present, or even most pronounced, in the most-difficult dual language context. By comparing this task with the bilingual switching task in Experiment 2, we aim to better understand whether these groups of older adults performed at a similar level as younger adults regardless of the task or whether an absence of age-group differences is specific to within-language switching.

### Task switching
Experiment 1 also included a (non-linguistic) task switching paradigm. Surprisingly, this task did not show a switching cost in terms of RTs and showed a switching benefit in terms of accuracy, contrary to a vast literature observing costs when switching between a range of tasks [51]. A large mixing cost was observed, showing that using different rules was more difficult than being in a single-rule condition. It is possible that the dual-rule condition was very difficult overall, even when not having to switch between rules. Even on non-switch trials, participants had to choose between one of two possible responses corresponding to the rule (for example, smaller or larger than 5 within the size-rule task). Responses in a linguistic noun-verb naming task might be more automatic, as language production is natural and people are fast at retrieving a word like “horse” in response to a picture. In contrast, determining whether 7 is smaller or larger than 5 might be difficult even on non-switch trials and is not a task people frequently perform in daily life. Furthermore, this might have been even more difficult in the online set-up, where participants had to click on, and remember, buttons corresponding to these options. Indeed, both age groups responded relatively slowly while doing this task. Older adults were significantly slower than younger adults. However, this larger age-group difference in terms of overall RTs could be explained by the general difficulty of the task and the absence of a strict time-out.

The very large mixing cost further confirms that the dual-rule context was difficult for both age groups and that participants might have put in much additional effort to respond to both non-switch and switch trials. A small difference in mixing costs between age groups was observed in the analysis on log RTs, although the direction varied for untransformed and log-transformed RTs. While the untransformed RTs suggest older adults had a larger mixing cost, the analysis on log RTs showed a smaller mixing cost for older adults. It is likely that the untransformed RTs show a larger raw mixing cost for older adults as their overall RTs were much slower than that of younger adults. This large increase in overall response times in older adults was only observed in the non-linguistic task, potentially also because the time out was much longer in this task (ten seconds). We therefore also conducted an additional analysis with z-scored RTs to cancel out the overall age effect (and indeed, in this analysis there was no overall RT age-group difference). This analysis no longer showed a significant age-group difference in terms of mixing costs. In combination with the benefit older adults showed in terms of the accuracy mixing cost too, it appears that older adults’ responses were not affected more negatively by dual-rule compared to single-rule contexts than younger adults’ responses. Thus, overall, neither the linguistic nor the non-linguistic tasks showed disadvantages for older adults.

@@

As a second aim, we examined how age affects within- and between-language control. Experiment 1 showed no clear language control differences between age groups, but previous literature has shown substantial age effects on bilingual control in terms of both reactive and proactive control mechanisms. If older adults show greater difficulty with proactive control, we expected their mixing costs to be larger than for younger adults. If these age-related changes are comparable across within- and between-language control, we expected these effects to be observed in both tasks, and for the age-group differences to be comparable. Similarly, if older adults show greater difficulty with reactive control, we expected their switching costs to be larger than those of younger adults, again also examining if age-related patterns are comparable for the within- and between-language control tasks.

@@

### Participants
Experiment 2 was completed by fifty older adults (*M*age = 67.20 years old, *SD* = 5.60, range = 60-82) and fifty younger adults (*M*age = 23.1, *SD* = 4.97, range = 18-35). Participants provided informed consent at the start of the study. Ethics approval was given by the ethics committee in the Department of Psychology, University of York. The study was conducted in accordance with the Declaration of Helsinki, including a pre-registration. Thirty-four of the older adults identified as female and 16 as male. Within the group of younger adults, 34 identified as female, 13 as male, and three chose one of the other options. Thirty-eight of the older adults and twenty-five of the younger adults reported having completed a graduate degree. An additional seven participants were tested but not included because their audio files were not uploaded or audible (N=5) or because they did not follow the task instructions (N=2). The sample size was determined based on power simulations. We first ran a simulation assessing power to detect the switching and mixing costs observed for the noun-verb task in Experiment 1. With 100 participants, power to detect these effects was over 95%. Given that Experiment 1 showed no age-group differences, we based the power in relation to age groups on the bilingual switching study with older and younger adults reported in [22]. Power to detect a mixing cost difference between age groups was over 80%. Power to detect a switching cost difference between age groups, based on effect sizes in [22], was slightly lower (>60%). However, based on practical considerations regarding recruitment feasibility, we decided to accept this power with 100 participants.

All participants reported that they were native speakers of English and met the same inclusion criteria as reported in Experiment 1. We again used the same Prolific screening settings and eligibility criteria regarding no diagnosis of a neurodegenerative disease or cognitive impairment. We did not use a separate task of cognitive functioning due to the study being completed online. Participants were bilingual in Experiment 2 and spoke French (39 older adults; 34 younger adults) or German (11 older adults; 16 younger adults) as a second language. We recruited both French- and German-speakers to facilitate recruitment but did not aim to compare French and German speakers. Participants completed the LexTALE as a short measure of vocabulary in each language ([52],[53]). In this task, they saw letter strings and had to indicate whether they formed an existing word in the target language or not. Participants also completed a language background questionnaire, which included questions about age of acquisition, self-rated proficiency, language use, and language switching. Participants reported being unbalanced bilinguals, with significantly higher proficiency in and use of English than their L2 (see Supplementary Table 6). Most participants were living in an English-dominant country (typically the UK). Some participants also reported speaking more than two languages, but for the purpose of our study, we always counted French or German as the L2, depending on the language version of the task they completed. We initially pre-registered to exclude participants who scored below 50% correct on the LexTALE. However, given that we aimed for unbalanced bilinguals with a lower L2 proficiency, several participants scored below this threshold and we removed this exclusion criterion.

@@

### Materials
Pictures for the two switching tasks were chosen in a similar way as in Experiment 1, with the list of stimuli available on the OSF page. We used the same pictures in the bilingual and in the noun-verb switching task to make them more comparable. Identical cognates between the languages were avoided, which also resulted in a different set of stimuli for the noun-verb task than in Experiment 1. For some pictures, multiple verbs were possible (e.g., “chew”, “brush”, or “bite”), which we all accepted. Words were matched in terms of number of phonemes between languages or nouns/verbs, but the verb and L2 frequency was significantly higher than the noun and L1 frequencies. Materials for the semantic control task were based on [7]. Again, two cues per switching task were used. The cues for the noun-verb switching task were the same as in Experiment 1. For the bilingual switching task, we could not create two written cues per language and we therefore worked with two visual cues in the form of two versions of the country flag associated with the language.

### Procedure
Participants were recruited through Prolific.co and through our existing database of older adult volunteers. They completed the study on Gorilla.sc. The procedure of the study was similar to Experiment 1, with the order of the noun-verb and bilingual switching tasks counterbalanced across participants. The tasks looked the same as in Experiment 1, with the change that each picture was presented for 3000ms. After the two switching tasks, participants completed a semantic control task based on [7]. In this task, participants saw a target word and two or four answer options. They had to choose the answer option that was similar in either size (e.g., donkey–bicycle) or colour (e.g., cloud-tooth). Other answer options were either neutral (congruent trials) or distractors (incongruent trials) that were related to the target in another way. For example, on the incongruent trial asking what is most similar to “gym” in size, participants had to choose “theatre” while ignoring the related word “basketball”. Half of the trials were incongruent while the other half were congruent. There was no time limit on this task. After completing these tasks, participants completed the LexTALE tasks as a measure of vocabulary and the language background questionnaire, as described in the Participants section.

@@

Mean accuracy per condition and language group is provided in Supplementary Table \@ref(tab:TableS7). The full results from the analysis are shown in Supplementary Table \@ref(tab:TableS8). The bilingual switching task showed a significant switching cost in terms of accuracy, with participants making more mistakes on switch than on non-switch trials. They were also more accurate in their L1 than in their L2. Accuracy did not show any effects of, or interactions with, age group.

@@

The full results of the switching-effect analysis can be found in Table \@ref(tab:Table4), with Table \@ref(tab:Table5) showing the means per condition. Participants showed a significant switching cost, with slower responses on switch trials (*M* = `r round(Bil_Trialtype_RT_M[3,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[3,2])`) than on non-switch trials (*M* = `r round(Bil_Trialtype_RT_M[1,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[1,2])`). Older adults (*M* = `r round(Bil_Age_RT_M[1,2])`, *SD* = `r round(Bil_Age_RT_SD[1,2])`) responded more slowly than the younger adults (*M* = `r round(Bil_Age_RT_M[2,2])`, *SD* = `r round(Bil_Age_RT_SD[2,2])`). Of main interest for the current study, older adults had a slightly higher switching cost than younger adults, but this did not reach significance (see Table \@ref(tab:Table4) and Figure 1). None of the other main effects or interactions were significant either.

@@

knitr::kable(table4data, booktabs = T, caption='Outcome of the linear mixed effect models for the bilingual switching task in Experiment 2. The final switching model included all by-subject and by-item random intercepts and slopes, after removal of correlations. The final mixing model included all by-subject and by-item random intercepts and slopes, after removal of correlations.', align='lrrrr') %>%
  row_spec(c(0,1,10),bold=T) %>%
  row_spec(c(1,10),italic=T)
```

@@

#plot nounverb mixing cost
NVMix = ggplot(means_NVMix_Exp2_long, aes(x = Agegroup, y = NounVerbMixing)) + geom_boxplot() +
  labs(title="D. Noun-verb mixing cost",x="Age group", y = "mixing cost (ms)")+
  theme(axis.text.x = element_text(color = "black", size = 12, face = "plain"),
        axis.text.y = element_text(color = "black", size = 12, face = "plain"),
        axis.title.x = element_text(color = "black", size = 14, face = "bold"),
        axis.title.y = element_text(color = "black", size = 14, face = "bold"),
        title = element_text(color = "black", size = 14, face = "bold.italic"), 
        strip.text.x = element_text(color = "black", size = 12, face = "plain"))+

@@

The full details are provided in Supplementary Tables \@ref(tab:TableS7) and \@ref(tab:TableS8). Participants showed a significant mixing cost, with higher accuracy on single-language trials than on dual-language trials. They were also more accurate in their L1 than L2. The mixing cost itself was also larger for the L1. The difference between languages (L1 naming advantage) was more pronounced on single-language than dual-language trials. Finally, this asymmetry in mixing cost interacted with age, as the difference between languages was largest for the younger adults.

Participants showed a significant RT mixing cost, with slower responses on non-switch trials (*M* = `r round(Bil_Trialtype_RT_M[1,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[1,2])`) than on single-language trials (*M* = `r round(Bil_Trialtype_RT_M[2,2])`, *SD* = `r round(Bil_Trialtype_RT_SD[2,2])`). Across the single-language and non-switch trials, older adults (M = 1229, SD = 167) also responded more slowly than younger adults (M = 1136, SD = 145). Of main interest for the current study, mixing costs were larger in the older adults than in the younger adults (see Table 5 & Figure 1).

@@

Given that older adults also responded significantly slower overall than younger adults, we z-scored the RTs to account for overall slowing. These analyses aimed to make sure that the larger mixing costs (and to some extent larger switching costs) for older adults were not the consequence of their overall slower naming. These analyses still showed a significant interaction between mixing costs and age ($\beta$ = `r round (MixingZ$coefficients[c(5),c(1)],digits=3)`, *SE* = `r round (MixingZ$coefficients[c(5),c(2)],digits=3)`, *t* = `r round (MixingZ$coefficients[c(5),c(4)],digits=3)`, *p* = `r p_format (MixingZ$coefficients[c(5),c(5)],digits=1,leading.zero=TRUE,accuracy=0.001)`), and the interaction between switching cost and age was now also significant ($\beta$ = `r round (SwitchingZ$coefficients[c(5),c(1)],digits=3)`, *SE* = `r round (SwitchingZ$coefficients[c(5),c(2)],digits=3)`, *t* = `r round (SwitchingZ$coefficients[c(5),c(4)],digits=3)`, *p* = `r p_format (SwitchingZ$coefficients[c(5),c(5)],digits=1,leading.zero=TRUE,accuracy=0.001)`).

@@

Mean accuracy by condition is shown in Supplementary Table \@ref(tab:tableS9), with full results from the analysis shown in Supplementary Table \@ref(tab:TableS10). There was a switch cost in terms of accuracy, with better performance on non-switch than switch trials. This interacted with rule, such that the switching cost was larger for the nouns. There were no effects of, or interactions with, age.
